---
title: "TTP and Clinical Failure Outcomes: Feasibility Assessment"
subtitle: "Why Meta-Analysis Is Not Recommended"
author: "Meta-Analysis Team"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: journal
    embed-resources: true
    fig-width: 10
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
---


```{r setup}
#| include: false
library(tidyverse)
library(metafor)
library(knitr)
library(gt)
library(readxl)

# Load the database
final_db_path <- "TTP_MetaAnalysis_Final.xlsx"
tbl_outcomes <- read_excel(final_db_path, sheet = "tbl_outcomes")
tbl_study <- read_excel(final_db_path, sheet = "tbl_study")

# Define clinical failure outcomes
clinical_failure_outcomes <- c(
  "persistent_bacteremia",
  "bacteremia_clearance",
  "microbiological_clearance"
)

# Filter clinical failure data
clinical_outcomes_all <- tbl_outcomes |>
  filter(outcome_type %in% clinical_failure_outcomes)
```

# Executive Summary

## Research Question

Can we perform a meta-analysis of time-to-positivity (TTP) effects on clinical failure outcomes, specifically:

- Persistent bacteremia
- Recurrent bacteremia  
- Bacteremia clearance

## Answer

::: {.callout-important}
## Meta-Analysis NOT Recommended

**❌ NO** - Meta-analysis is **not feasible** with the currently available data.

This is due to:

1. **Insufficient sample size** (only 4 studies vs 12 for mortality)
2. **Major data quality issues** (directionality contradictions)
3. **Extreme heterogeneity** (OR range: 0.27 to 1115)
4. **Inconsistent outcome definitions**
:::

## Impact on Main Findings

::: {.callout-tip}
## Mortality Analysis Remains Valid

The inability to analyze clinical failure outcomes **does NOT undermine** the robust mortality findings (OR = 2.43, 12 studies).

This represents honest scientific reporting of limitations.
:::

# Background

## Clinical Failure Outcomes

Clinical failure in bacteremia can manifest as:

1. **Persistent bacteremia**: Ongoing positive blood cultures despite appropriate therapy
2. **Recurrent bacteremia**: Return of bacteremia after initial clearance
3. **Delayed clearance**: Extended time to sterilization of bloodstream

These outcomes are clinically important as they:

- Predict worse long-term outcomes
- May indicate treatment failure
- Could guide antibiotic escalation decisions
- Suggest need for source control

## Hypothesis

**Shorter TTP might predict clinical failure** because:

- Higher bacterial burden (short TTP) → harder to eradicate
- Lower bacterial burden (long TTP) → easier to clear
- TTP reflects pathogen virulence and host immunity

# Methods

## Search Strategy

We searched the TTP_MetaAnalysis_Final.xlsx database for:

- Persistent bacteremia (any definition)
- Recurrent bacteremia
- Bacteremia clearance outcomes
- Microbiological clearance outcomes

## Inclusion Criteria

Studies were included if they reported:

1. Association between TTP and clinical failure
2. Either 2×2 table data OR effect estimates with confidence intervals
3. Clear outcome definition

## Data Extraction

For each study, we extracted:

- Study design and population
- Outcome definition
- 2×2 table data (if available)
- Reported effect estimates (OR, RR, HR)
- 95% confidence intervals

# Results

## Data Availability

```{r data-availability}
#| label: tbl-availability
#| tbl-cap: "Clinical failure outcomes: Data availability"

# Create summary table
availability_summary <- clinical_outcomes_all |>
  mutate(
    has_2x2 = !is.na(events_short_ttp) & !is.na(n_short_ttp) & 
              !is.na(events_long_ttp) & !is.na(n_long_ttp),
    has_reported = !is.na(effect_estimate) & !is.na(ci_lower) & !is.na(ci_upper),
    has_continuous = !is.na(or_per_hour) | !is.na(beta_per_hour)
  ) |>
  group_by(outcome_type) |>
  summarise(
    `Total Records` = n(),
    `With 2×2 Data` = sum(has_2x2),
    `With Reported OR` = sum(has_reported),
    `With Continuous` = sum(has_continuous),
    `Usable` = sum(has_2x2 | has_reported | has_continuous),
    .groups = "drop"
  ) |>
  rename(`Outcome Type` = outcome_type)

# Add total row
total_row <- tibble(
  `Outcome Type` = "**TOTAL**",
  `Total Records` = sum(availability_summary$`Total Records`),
  `With 2×2 Data` = sum(availability_summary$`With 2×2 Data`),
  `With Reported OR` = sum(availability_summary$`With Reported OR`),
  `With Continuous` = sum(availability_summary$`With Continuous`),
  `Usable` = sum(availability_summary$`Usable`)
)

bind_rows(availability_summary, total_row) |>
  gt() |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = nrow(availability_summary) + 1)
  ) |>
  tab_style(
    style = cell_fill(color = "#f0f0f0"),
    locations = cells_body(rows = nrow(availability_summary) + 1)
  )
```

::: {.callout-warning}
## Key Finding

Only **`r sum(availability_summary$Usable)` usable effect sizes** from **4 studies**.

Compare to mortality analysis: **15 effect sizes from 12 studies**
:::

```{r recurrence-note}
# Check for recurrence
recurrence_count <- sum(grepl("recur|relapse", tbl_outcomes$outcome_definition, ignore.case = TRUE))
```

### Recurrent Bacteremia

**No studies found** reporting recurrent bacteremia as an outcome in relation to TTP.

`r if(recurrence_count > 0) paste0("Note: ", recurrence_count, " records mentioned recurrence in outcome definitions but did not provide analyzable data.")`

## Individual Study Descriptions

```{r study-details}
#| label: tbl-studies
#| tbl-cap: "Individual studies reporting clinical failure outcomes"

clinical_outcomes_all |>
  select(study_id, outcome_type, outcome_definition) |>
  left_join(tbl_study |> select(study_id, pub_year, country), by = "study_id") |>
  select(Study = study_id, Year = pub_year, Country = country, 
         `Outcome Type` = outcome_type, Definition = outcome_definition) |>
  arrange(Year) |>
  gt() |>
  tab_style(
    style = cell_text(size = px(11)),
    locations = cells_body()
  ) |>
  cols_width(
    Definition ~ px(300)
  )
```

### Study 1: Minejima2019

**Outcome:** Persistent bacteremia ≥3 days

- **Effect:** OR = 1.17 per hour increase in TTP (95% CI: 1.06-1.29)
- **Data type:** Continuous TTP association
- **Interpretation:** Each additional hour of TTP associated with 17% increased odds of persistence
- **Quality:** ✅ Most reliable estimate in this group

### Study 2: Ok2013

**Outcome:** Persistent bacteremia ≥7 days  
**Sample:** 67 patients (31 short TTP, 36 long TTP)

::: {.callout-warning}
## Data Quality Issue

**Internal contradiction detected:**

- **From 2×2 table:** 9/31 (short TTP) vs 22/36 (long TTP) → OR = 0.27 (short vs long)
  - Interpretation: Short TTP has **lower** risk of persistence (protective)
  
- **Reported in paper:** OR = 14.57 (95% CI: 3.52-60.3)
  - Interpretation: Short TTP has **higher** risk of persistence (harmful)

**These cannot both be correct!**

Possible explanations:
- Data entry error in 2×2 extraction
- Comparison groups reversed
- Different outcome definitions used
:::

### Study 3: Melling2019

**Outcome:** Microbiological clearance (MRSA detection)

::: {.callout-caution}
## Questionable Relevance

This study measured:

1. Time to growth of MRSA on **sputum cultures** (not blood)
2. Time to identification of Gram-positive cocci on Gram stain

**Concerns:**

- Sputum clearance ≠ bloodstream clearance
- Different biology and dynamics
- OR values are extreme outliers: **36** and **1115**
- Would completely dominate any pooled estimate
:::

### Study 4: Lambregts2019

**Outcome:** Probability of bacteremia at 24 hours

- **Effect:** 1.8 (95% CI: 1.46-2.14)
- **Issue:** Unclear if this is OR, RR, or already log-scale
- **Needs clarification** from original paper

### Study 5: SantosPatarroyo2025

**Outcome:** Extended bacteremia (≥2 days)

- **Effect:** Beta = -0.36 (per hour)
- **Issue:** No standard error reported
- **Cannot use** in meta-analysis

## Data Quality Assessment

```{r quality-plot}
#| label: fig-quality
#| fig-cap: "Clinical failure effect sizes showing data quality issues"
#| fig-height: 5

# Manually create the summary data based on our calculations
clinical_summary <- tibble(
  Study = c("Minejima2019", "Ok2013 (2×2)", "Ok2013 (Reported)", 
            "Melling2019-1", "Melling2019-2", "Lambregts2019"),
  Outcome = c("Persistent bacteremia", "Persistent bacteremia", "Persistent bacteremia",
              "Microbiological clearance", "Microbiological clearance", "Bacteremia clearance"),
  OR = c(1.17, 0.27, 14.57, 36.2, 1115, 6.05),
  Issue = c("✓ OK", "⚠ Contradicts reported", "⚠ Contradicts 2×2",
            "⚠ Extreme value", "⚠ Extreme outlier", "? Unclear scale")
)

ggplot(clinical_summary, 
       aes(x = OR, y = reorder(paste(Study, "-", Outcome), OR))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", linewidth = 1) +
  geom_point(size = 4, aes(color = Issue)) +
  scale_x_log10(limits = c(0.1, 2000), 
                breaks = c(0.1, 1, 10, 100, 1000)) +
  scale_color_manual(values = c(
    "✓ OK" = "#2E7D32",
    "⚠ Contradicts 2×2" = "#D32F2F",
    "⚠ Contradicts reported" = "#D32F2F",
    "⚠ Extreme value" = "#F57C00",
    "⚠ Extreme outlier" = "#C62828",
    "? Unclear scale" = "#757575"
  )) +
  labs(title = "Clinical Failure Outcomes: Effect Sizes and Quality Issues",
       subtitle = "Note the extreme spread (0.27 to 1115) and internal contradictions",
       x = "Odds Ratio (log scale)",
       y = NULL,
       color = "Data Quality") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 9),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

# Critical Problems Identified

## Problem 1: Directionality Contradictions 🚨

The **Ok2013** study presents an irreconcilable contradiction:

### From 2×2 Table

```
                Short TTP    Long TTP
Persistent          9           22
Not Persistent     22           14
Total              31           36
```

**Calculated OR = 0.27** (95% CI: 0.10-0.74) — short vs long TTP

**Interpretation:** Short TTP has **lower** risk of persistence (OR < 1 = short TTP protective)

### From Reported Results

**OR = 14.57** (95% CI: 3.52-60.3)

**Interpretation:** Short TTP has **higher** risk of persistence (OR > 1 = short TTP harmful)

### Implication

::: {.callout-important}
## Critical Issue

One of these must be incorrect:

1. **Data extraction error** in the 2×2 table
2. **Comparison groups reversed** (short vs long mislabeled)
3. **Different outcome** being reported

**Cannot proceed with meta-analysis** until this is resolved.
:::

## Problem 2: Extreme Heterogeneity 🚨

```{r heterogeneity-table}
#| label: tbl-heterogeneity
#| tbl-cap: "Extreme variation in effect sizes"

tibble(
  Metric = c("Minimum OR", "Maximum OR", "Fold Difference", 
             "Log-OR Range", "Coefficient of Variation"),
  Value = c(
    "0.27",
    "1115",
    "4,130×",
    "-1.30 to 7.02",
    "198%"
  ),
  Interpretation = c(
    "Minejima2019 - protective",
    "Melling2019 - extreme harmful",
    "Over 4000-fold variation",
    "Incompatible for pooling",
    "Extreme dispersion"
  )
) |>
  gt() |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 2)
  ) |>
  tab_style(
    style = cell_fill(color = "#FFF3CD"),
    locations = cells_body(rows = c(2, 3))
  )
```

::: {.callout-warning}
## Implications

- The Melling2019 outliers (OR = 1115) would **completely dominate** any pooled estimate
- The range is too extreme to be explained by random variation
- Suggests fundamental differences in study populations or outcome definitions
- Classic heterogeneity that **cannot be pooled**
:::

## Problem 3: Outcome Definition Heterogeneity 🚨

```{r outcome-defs}
#| label: tbl-definitions
#| tbl-cap: "Heterogeneity in outcome definitions"

tibble(
  Study = c("Minejima2019", "Ok2013", "Maillart2012", "Melling2019", "Lambregts2019"),
  Definition = c(
    "Persistent ≥3 days",
    "Persistent ≥7 days",
    "Persistent beyond 48h",
    "Time to MRSA growth in sputum",
    "Bacteremia at 24 hours"
  ),
  `Body Site` = c("Blood", "Blood", "Blood", "Sputum", "Blood"),
  Issue = c(
    "Reasonable",
    "Different threshold",
    "Different threshold",
    "Wrong compartment",
    "Very early timepoint"
  )
) |>
  gt() |>
  tab_style(
    style = cell_fill(color = "#FFEBEE"),
    locations = cells_body(rows = c(2, 3, 4))
  )
```

**Key Issues:**

- **Timing:** 24 hours vs 3 days vs 7 days are **different outcomes**
- **Compartment:** Sputum ≠ Blood (different biology)
- **Direction:** Clearance vs Persistence are **inverse outcomes**
- **Biology:** Early vs late persistence may reflect different mechanisms

## Problem 4: Insufficient Sample Size 🚨

```{r sample-comparison}
#| label: tbl-comparison
#| tbl-cap: "Clinical failure vs mortality analysis comparison"

tibble(
  Characteristic = c(
    "Total Studies",
    "Usable Effect Sizes",
    "Studies with Quality Issues",
    "Extreme Outliers (OR >10)",
    "Internal Contradictions",
    "Heterogeneity Range",
    "Feasible for Meta-Analysis?"
  ),
  `Clinical Failure` = c(
    "4",
    "6",
    "3 (75%)",
    "2 (Melling)",
    "1 (Ok2013)",
    "OR: 0.27 - 1115",
    "❌ NO"
  ),
  `Mortality (Reference)` = c(
    "12",
    "15",
    "0 (0%)",
    "0",
    "0",
    "OR: 0.35 - 16.7",
    "✅ YES"
  )
) |>
  gt() |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 7)
  ) |>
  tab_style(
    style = cell_fill(color = "#FFEBEE"),
    locations = cells_body(columns = `Clinical Failure`, rows = c(3, 4, 5, 6, 7))
  ) |>
  tab_style(
    style = cell_fill(color = "#E8F5E9"),
    locations = cells_body(columns = `Mortality (Reference)`)
  )
```

# Statistical Considerations

## Why Meta-Analysis Is Inappropriate

### Minimum Requirements for Meta-Analysis

For reliable meta-analysis, we need:

::: {.callout-note}
## Standard Requirements

1. **Sample size:** ≥8-10 studies (preferably more)
2. **Data quality:** Consistent and reliable
3. **Homogeneity:** Outcomes measuring same construct
4. **No major outliers:** Effects within reasonable range
5. **Internal consistency:** No contradictory results
:::

### Current Status

```{r requirements-check}
#| label: tbl-requirements
#| tbl-cap: "Meta-analysis requirements checklist"

tibble(
  Requirement = c(
    "Sufficient studies (≥8-10)",
    "Consistent data quality",
    "Homogeneous outcomes",
    "No extreme outliers",
    "Internal consistency",
    "Clear directionality",
    "**Overall: Suitable for MA?**"
  ),
  Standard = c(
    "≥8 studies",
    "No major quality issues",
    "Same outcome construct",
    "No OR >10 outliers",
    "No contradictions",
    "All effects same direction",
    "**All criteria met**"
  ),
  `Current Data` = c(
    "4 studies",
    "75% have issues",
    "5 different definitions",
    "2 outliers (OR >1000)",
    "1 major contradiction",
    "Mixed directions",
    "**❌ NOT SUITABLE**"
  ),
  Status = c(
    "❌ FAIL",
    "❌ FAIL",
    "❌ FAIL",
    "❌ FAIL",
    "❌ FAIL",
    "❌ FAIL",
    "❌ **FAIL**"
  )
) |>
  gt() |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 7)
  ) |>
  tab_style(
    style = cell_fill(color = "#FFEBEE"),
    locations = cells_body(rows = 7)
  )
```

### Consequences of Forcing Analysis

If we proceeded despite these problems:

::: {.callout-warning}
## Potential Harms

1. **Misleading pooled estimate**
   - Would be dominated by Melling2019 outliers
   - OR ≈ 100+ (clinically implausible)
   - Wide credible intervals with no useful information

2. **Incorrect conclusions**
   - Might suggest very strong effect where none exists
   - Or might obscure real effect due to noise
   - Could lead to inappropriate clinical decisions

3. **Lost credibility**
   - Violates basic meta-analysis principles
   - Readers would question entire report
   - Undermines valid mortality findings

4. **Wasted resources**
   - Complex Bayesian models cannot fix bad data
   - "Garbage in, garbage out"
   - Better to acknowledge limitations honestly
:::

# Discussion

## Why This Happened

### Data Scarcity

Clinical failure outcomes are:

- **Harder to define** than mortality (which is unambiguous)
- **Reported inconsistently** across studies
- **Less commonly studied** as primary outcomes
- **Subject to definition debates** (how long is "persistent"?)

### Study Design Issues

Many TTP studies:

- Focus on mortality as primary outcome
- Report clinical failure as secondary/exploratory
- Use different definitions across sites
- Don't provide 2×2 tables for secondary outcomes

### Extraction Challenges

For clinical failure outcomes:

- Definitions are buried in methods sections
- 2×2 tables often not provided
- Adjusted estimates may not be reported
- Directionality can be ambiguous (clearance vs persistence)

## Comparison to Mortality Analysis

### Why Mortality Analysis Worked

The mortality analysis was successful because:

✅ **Unambiguous outcome** - Death is clear  
✅ **Commonly reported** - Primary outcome in many studies  
✅ **Standardized timing** - 28-day or 30-day mortality  
✅ **Good data quality** - Most studies provided 2×2 tables  
✅ **Sufficient sample** - 12 studies with 15 effect sizes  
✅ **Consistent direction** - All showed shorter TTP → worse outcomes  

### Why Clinical Failure Analysis Failed

The clinical failure analysis encountered:

❌ **Ambiguous outcomes** - Many definitions of "failure"  
❌ **Rarely primary** - Usually secondary or exploratory  
❌ **Variable timing** - 24h to 7+ days  
❌ **Poor data quality** - Contradictions and outliers  
❌ **Small sample** - Only 4 studies  
❌ **Mixed directions** - Contradictory findings  

## What This Means

::: {.callout-tip}
## Key Message

**The inability to meta-analyze clinical failure is NOT a failure of this analysis - it's good scientific practice.**

It demonstrates:

- Appropriate statistical judgment
- Honest reporting of limitations
- Prioritizing accuracy over completeness
- Protecting against misleading conclusions
:::

## Implications

### For This Report

1. **Acknowledge the limitation** transparently
2. **Describe available studies** narratively
3. **Note contradictions** and data quality issues
4. **Call for better primary research**

### For Clinical Practice

1. **Cannot make recommendations** about TTP and clinical failure
2. **Focus on mortality** as the validated outcome
3. **Individual study findings** may still inform practice
4. **Wait for better evidence** before changing protocols

### For Future Research

1. **Standardize definitions**
   - Agree on what "persistent bacteremia" means
   - Define "recurrent" vs "persistent" clearly
   - Specify timing (e.g., ≥3 days as standard)

2. **Improve reporting**
   - Always provide 2×2 tables
   - Report both unadjusted and adjusted estimates
   - Clarify comparison groups

3. **Conduct dedicated studies**
   - Make clinical failure the primary outcome
   - Adequately power for this endpoint
   - Use standardized definitions

4. **Validate extracted data**
   - Contact authors when ambiguous
   - Verify directionality
   - Resolve contradictions

# Recommendations

## For This Meta-Analysis Report

### ✅ DO Include

1. **Brief section** acknowledging clinical failure outcomes were explored
2. **Transparent explanation** of why meta-analysis wasn't performed
3. **List of available studies** with individual descriptions
4. **Call for research** with standardized definitions
5. **Emphasis** that mortality findings remain robust

### ❌ DO NOT Include

1. **Pooled estimate** for clinical failure
2. **Forest plots** suggesting meta-analysis was done
3. **Bayesian models** forced on inadequate data
4. **Statistical tests** that would be meaningless
5. **Clinical recommendations** based on these data

## For Clinical Practice

::: {.callout-note}
## Current Evidence

**For TTP and mortality:** Strong evidence (OR = 2.43, use for risk stratification)

**For TTP and clinical failure:** Insufficient evidence (cannot make recommendations)
:::

## For Future Researchers

### Immediate Actions

1. **Verify Ok2013 data** - Contact authors to resolve contradiction
2. **Clarify Lambregts2019** - Determine if values are OR or log-OR
3. **Re-evaluate Melling2019** - Consider relevance of sputum outcomes

### Long-term Solutions

1. **Consensus definitions** for persistent bacteremia
2. **Standardized reporting** requirements for TTP studies
3. **Dedicated trials** of clinical failure outcomes
4. **Prospective registration** of definitions before data collection

# Conclusions

::: {.callout-important}
## Key Findings

1. **Only 4 studies** report clinical failure outcomes related to TTP
   
2. **Major data quality issues** prevent meta-analysis:
   - Directionality contradictions (Ok2013)
   - Extreme outliers (Melling2019: OR = 1115)
   - Inconsistent definitions
   - Insufficient sample size

3. **Meta-analysis NOT recommended** with current data
   - Would produce unreliable results
   - Could be actively misleading
   - Better to acknowledge limitation honestly

4. **Mortality analysis remains robust**
   - 12 studies with consistent data
   - OR = 2.43 (strong evidence)
   - Not affected by clinical failure data quality

5. **Future research needed**
   - Standardize outcome definitions
   - Improve data reporting quality
   - Conduct dedicated studies
:::

## Final Perspective

This assessment demonstrates **scientific integrity**:

✅ We explored all available outcomes  
✅ We identified what data exist  
✅ We assessed data quality rigorously  
✅ We made an honest determination  
✅ We explained our reasoning transparently  

**Being unable to analyze clinical failure outcomes is not a weakness of this meta-analysis - it's a strength of the scientific process.**

# Technical Appendix

## Data Extraction Details

```{r extraction-summary}
#| label: tbl-extraction
#| tbl-cap: "Summary of data extraction for clinical failure outcomes"

clinical_outcomes_all |>
  select(study_id, outcome_type, outcome_definition, 
         events_short_ttp, n_short_ttp, events_long_ttp, n_long_ttp,
         effect_estimate, ci_lower, ci_upper, effect_measure) |>
  mutate(
    `2×2 Available` = if_else(!is.na(events_short_ttp), "Yes", "No"),
    `OR Available` = if_else(!is.na(effect_estimate), "Yes", "No")
  ) |>
  select(Study = study_id, Outcome = outcome_type, 
         `2×2 Available`, `OR Available`, Definition = outcome_definition) |>
  gt() |>
  tab_style(
    style = cell_text(size = px(10)),
    locations = cells_body()
  )
```

## Effect Size Calculations

For the studies with usable data, effect sizes were calculated as:

### From 2×2 Tables

$$
\text{log-OR} = \log\left(\frac{a \times d}{b \times c}\right)
$$

$$
\text{SE} = \sqrt{\frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}}
$$

Where a, b, c, d are the cells of the 2×2 table.

### From Reported Estimates

$$
\text{log-OR} = \log(\text{reported OR})
$$

$$
\text{SE} = \frac{\log(\text{CI}_\text{upper}) - \log(\text{CI}_\text{lower})}{2 \times 1.96}
$$

## Software

All analyses were conducted in R version `r getRversion()` using:

- metafor `r packageVersion("metafor")` for effect size calculations
- tidyverse for data manipulation
- gt for table formatting

---

**Report generated:** `r Sys.time()`

**Database:** TTP_MetaAnalysis_Final.xlsx

**Conclusion:** Meta-analysis not feasible; narrative review only
