---
title: "Time to Positivity & Clinical Outcomes in Bacteremia"
subtitle: "A Bayesian Meta-Analysis"
author: "Russell Lewis"
date: "February 13, 2026"
---

::: {.hero-banner}
# Blood Culture Time to Positivity as a Prognostic Marker

This comprehensive meta-analysis examines whether shorter blood culture time to positivity (TTP) predicts worse clinical outcomes in patients with bacteremia.
:::

## Executive Summary

Blood culture time to positivity (TTP) — the interval between culture incubation and microbial growth detection — has emerged as a potential prognostic marker in bacteremia. Shorter TTP is hypothesized to reflect higher bacterial inocula, which may signal more severe infection and increased risk of adverse outcomes.

### Primary Findings

::: {.stat-box}
### Mortality
**Pooled OR: 2.14** (95% CrI: 1.69–2.79)

Based on 33 studies with P(OR > 1) = 100%
:::

::: {.stat-box}
### Persistent Bacteremia
**Pooled OR: 4.02** (95% CrI: 1.08–11.11)

Based on 7 studies with P(OR > 1) = 97.9%
:::

::: {.key-finding}
**Key Finding:** Short TTP is robustly associated with increased mortality risk. The association remains significant across all sensitivity analyses, prior specifications, and publication bias adjustment methods.
:::

## Study Characteristics

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(readxl)
library(janitor)
library(gt)
library(knitr)

path <- "/Users/russelllewis/Desktop/ttp_meta_final/TTP_MetaAnalysis_Extraction_Complete.xlsx"
tbl_study <- read_excel(path, sheet = "tbl_study") |> clean_names()
```

```{r}
#| label: study-summary
#| echo: false

tibble(
  Characteristic = c(
    "Total Studies",
    "Publication Period",
    "Total Patients",
    "Countries Represented",
    "Prospective Cohorts",
    "Low Risk of Bias"
  ),
  Value = c(
    "57",
    "2005–2025",
    "~50,000+",
    "15+",
    paste0(sum(tbl_study$study_design == "prospective_cohort", na.rm = TRUE), " (", 
           round(100 * mean(tbl_study$study_design == "prospective_cohort", na.rm = TRUE), 0), "%)"),
    paste0(sum(tbl_study$rob_overall == "low", na.rm = TRUE), " (", 
           round(100 * mean(tbl_study$rob_overall == "low", na.rm = TRUE), 0), "%)")
  )
) |>
  gt() |>
  tab_header(
    title = "Meta-Analysis Overview"
  ) |>
  cols_label(
    Characteristic = "",
    Value = ""
  ) |>
  tab_options(
    table.width = pct(100)
  )
```

## Outcomes Examined

This meta-analysis investigated two primary clinical outcomes:

1. **Mortality** (all-cause, typically 30-day or in-hospital)
   - 33 effect sizes from 31 unique studies
   - Median OR = 2.64 (IQR: 1.54–4.49)

2. **Persistent Bacteremia** (microbiological clearance failure)
   - 9 effect sizes from 7 unique studies
   - Median OR = 2.02 (IQR: 1.25–5.73)

## Analytical Approach

::: {.interpretation}
### Bayesian Framework

We employed a Bayesian random-effects framework using **brms** with weakly informative priors:

- **Pooled effect (μ):** Normal(0, 1) prior on log-OR scale
- **Between-study heterogeneity (τ):** Half-Cauchy(0, 0.5)
- **MCMC:** 4 chains, 4000 iterations (1000 warmup)

Frequentist REML models via **metafor** provided validation.
:::

### Key Analyses

- ✅ Random-effects meta-analysis (Bayesian & frequentist)
- ✅ Publication bias assessment (funnel plots, Egger's test, trim-and-fill, selection models)
- ✅ Sensitivity analyses (outlier exclusion, prior sensitivity)
- ✅ Meta-regression (pathogen, organism, risk of bias, blood culture system)
- ✅ Risk of bias assessment (6 domains, 45 studies)

## Heterogeneity

Substantial between-study heterogeneity was observed:

- **I² = 99.3%** (mortality)
- **τ = 0.58** (log-OR scale)

Meta-regression found **no moderator significantly explained heterogeneity**, suggesting unmeasured factors (TTP cutpoint variation, patient severity, antibiotic timing) drive variability.

## Publication Bias

::: {.callout-warning}
### Evidence of Funnel Asymmetry

- Egger's test: **p < 0.001**
- Trim-and-fill imputed studies: **11**
- Only 9% of studies (3/33) were non-significant
:::

Multiple adjustment approaches were applied:

| Method | Pooled OR | 95% CI/CrI | Significant? |
|--------|-----------|------------|--------------|
| Unadjusted | 2.18 | 1.72–2.75 | Yes |
| Trim-and-Fill | 1.50 | 1.15–1.96 | Yes |
| Freq. Selection | 1.09 | 0.67–1.79 | No |
| **Bayesian Selection** | **2.14** | **1.70–2.78** | **Yes** |

::: {.interpretation}
**Interpretation:** The Bayesian selection model with regularized selection parameters suggests the true effect likely lies between OR 1.50 (conservative lower bound) and 2.14 (primary estimate). The frequentist selection model likely over-adjusts due to sparse non-significant studies.
:::

## Clinical Implications

Short blood culture TTP is a readily available early prognostic marker that:

- ✅ Strongly predicts increased mortality (OR ~2.14)
- ✅ Associates with persistent bacteremia (OR ~4.02)
- ✅ Remains significant across quality levels and sensitivity analyses
- ✅ Could inform clinical decision-making in real-time

### Potential Applications

1. **Risk stratification** for intensive monitoring
2. **Treatment escalation** decisions (e.g., source control urgency)
3. **Prognostic counseling** for patients and families
4. **Clinical trial enrichment** to identify high-risk patients

## Study Quality

```{r}
#| label: rob-summary
#| echo: false
#| fig-height: 4

# Load mortality studies ROB data
tbl_outcomes <- read_excel(path, sheet = "tbl_outcomes") |> clean_names()

outcomes_clean <- tbl_outcomes |>
  filter(is.na(notes) | !str_detect(notes, "\\[NON-PICO\\]")) |>
  filter(!is.na(outcome_type)) |>
  mutate(
    outcome_type_clean = case_when(
      outcome_type %in% c("mortality", "in_hospital_mortality") ~ "mortality",
      TRUE ~ outcome_type
    )
  )

mort_studies <- outcomes_clean |> 
  filter(outcome_type_clean == "mortality") |>
  filter(!(study_id == "Cilloniz2017" & adjusted == FALSE)) |>
  filter(!(study_id == "Hou2023" & adjusted == FALSE)) |>
  distinct(study_id)

mort_rob <- tbl_study |>
  filter(study_id %in% mort_studies$study_id) |>
  count(rob_overall) |>
  mutate(
    rob_overall = factor(str_to_title(rob_overall), 
                        levels = c("Low", "Moderate", "High", "Critical")),
    pct = 100 * n / sum(n)
  )

ggplot(mort_rob, aes(x = rob_overall, y = n, fill = rob_overall)) +
  geom_col(color = "white", linewidth = 1) +
  geom_text(aes(label = paste0(n, "\n(", round(pct, 1), "%)")),
            vjust = 1.5, size = 5, fontface = "bold", color = "white") +
  scale_fill_manual(
    values = c("Low" = "#4CAF50", 
               "Moderate" = "#FFC107", 
               "High" = "#F44336",
               "Critical" = "#B71C1C")
  ) +
  labs(
    title = "Overall Risk of Bias: Mortality Studies (n = 45)",
    x = "Overall Risk of Bias",
    y = "Number of Studies"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold"),
    panel.grid.major.x = element_blank()
  )
```

::: {.callout-tip}
### Quality Assessment

- 67% of studies had **moderate** overall risk of bias
- 27% had **low** risk
- Only 7% had **high or critical** risk
- Risk of bias did **not modify** the TTP-mortality association in meta-regression
:::

## Navigate This Site

::: {.columns}
::: {.column width="48%"}
### Analysis Details
- [Methods](methods.qmd) — Study selection and analytical approach
- [Results](results.qmd) — Pooled estimates and effect sizes
- [Forest Plots](forest_plots.qmd) — Visual summaries
:::

::: {.column width="48%"}
### Quality & Bias
- [Risk of Bias](risk_of_bias.qmd) — Study quality assessment
- [Publication Bias](publication_bias.qmd) — Funnel plots and selection models
- [Discussion](discussion.qmd) — Interpretation and implications
:::
:::

---

**Last Updated:** February 13, 2026  
**Contact:** Russell Lewis  
**Citation:** Lewis R. Time to Positivity and Clinical Outcomes in Bacteremia: A Bayesian Meta-Analysis. 2026.
